Synthesizing Symmetric Lenses

### 2018-07-10 papers/SymmetricLenses/main.tex
in progress edits to intro

While lenses facilitate writing principled data
transformations, \textit{synthesizing} lenses 
can make the task even easier.  Lens languages are a ripe target for
synthesis because lens languages, like many domain-specific languages,
reduce the space of possible programs, making search-based synthesis easier. 
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician, maina+:quotient-synthesis} has exploited this
observation by demonstrating that it is possible to synthesize relatively large,
complex lenses from specifications. In particular, given types that describe
source and target formats and optional examples of corresponding data, the
bijective synthesis algorithm presented in these 
papers will find bijective lenses to convert between data
% ~\cite{optician}
in ad hoc formats, such as linux 
configuration files that represent the same data in different ways. 
Often, these algorithms can find an appropriate lens
in less than a second without requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are simply not very
many different bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to be the one the
user wants.

@@

A natural way to expand the scope of such tools is to consider a more
powerful language of lenses.  In particular, \emph{symmetric
lenses}~\cite{symmetric-lenses} encompass a broad collection of bidirectional
transformations and include bijections as well as transformations that
allow each side of the transaction to retain information not present
in the other side.  If $\ell$ is a symmetric lens between formats $S$ and $T$, one can
use it to transform data from $S$ to $T$, or vice versa equally well.  A user would typically
run $\ell$ in the forwards direction (from $S$ to $T$) on a string $s$ when a user updates
$s$---executing $\ell$ in this direction will propagate those updates to $T$.
However, because data in format $T$ may contain some information not present in format $T$,
the $S \rightarrow T$ transformation will take
both a string $s \in S$ and a string $t \in T$ as arguments.
Inuitively, the lens should extract the data ``shared'' between the two formats from $s$
as it generates an output $t'$.  If $s$ is missing information, the transformation should
extract the missing components from $t$.

Initially, we hypothesized that extending the synthesis algorithms
of Miltner \ETAL~\cite{optician} to operate over symmetric lenses
as opposed to bijective lenses 
would be relatively straightforward:  Simply replace one set of combinators with another
and execute a search with similar heuristics.  However, in our early,
naive attempts to do so, we ran
into two different kinds of difficulties. \dpw{I got into the weeds a little bit there.
The goal was to set up the idea that there are ``two parts'' to a symmetric
lens transformation: the information from $s$ and the information from $t$.  In
a future paragraph in the intro we are going to explain that the central heuristic
involves minimizing the information from $t$.}

>>>>

While lenses facilitate writing principled data
transformations, \textit{synthesizing} lenses 
can make the task even easier.  Lens languages are a ripe target for
synthesis because lens languages, like many domain-specific languages,
reduce the space of possible programs, making search-based synthesis easier. 
%This feature can be a negative, if the desired transformation
%falls outside of the scope provided by the language, but can also be positive, as complex
%properties (like invertability) can be enforced automatically.
%Another advantage of reducing the
%set of programs is that doing so can facilitate program synthesis, which is essentially
%a search through the space of possible computations that can be expressed as a lens.
Recent research~\cite{optician, maina+:quotient-synthesis} has exploited this
observation by demonstrating that it is possible to synthesize relatively large,
complex lenses from specifications. In particular, given types that describe
source and target formats and optional examples of corresponding data, the
bijective synthesis algorithm presented in these 
papers will find bijective lenses to convert between data stored
% ~\cite{optician}
in different formats,
such as linux configuration files from different systems that represent the same data in different ways.  
Often, these algorithms can find an appropriate lens
in less than a second without requiring any examples.
%, with just types describing the source and target formats.
% \dpw{I would cite the ICFP paper (cite string:
%   maina+:quotient-synthesis) as to appear but I'm not sure how to do that
%   without possibly violating the spirit of double-blind reviewing. ie, if it
%   hasn't appeared yet, how could we know the content if we aren't the authors?
%   Mentioning it in the 3rd person does not seem to suffice ...} \ksf{The ICFP
%   2018 website lists the paper; I think it is okay to cite it. The reviewer can
%   assume we contacted the authors for a copy. We could ask Stephanie for a
%   ruling if we want to be sure. One of the principles of double blind as adopted
%   by SIGPLAN is not to reduce the quality of the paper; omitting this citation
%   would have that affect.}\bcp{+1}\saz{Stephanie says we should cite it in the
%   third person, even if that's a bit awkward, but otherwise citing it is OK.}
One reason that bijective lens synthesis is so effective, even relative to successful
synthesis projects in other domains, is that there are not very
many bijections between related data formats (unless the
same kind of data appears in the two formats many times). 
Hence, if the synthesis algorithm can
find any bijection at all, then it is very likely to be the one the
user wants.

@@

A natural way to expand the scope of such tools is to consider a more
exrpressive language of lenses.  For example, \emph{symmetric
lenses}~\cite{symmetric-lenses} encompass a broad collection of bidirectional
transformations that include bijections as well as more flexibile transformations that
allow each side to retain information not present
in the other.  If $\ell$ is a symmetric lens between formats $S$ and
$T$, then $\ell$  can transform data from $S$ to $T$ or vice versa
equally well.  A user would typically 
run $\ell$ in the forwards direction (from $S$ to $T$) when a user updates
a string $s \in S$ to propagate those updates to $T$.
However, because data in format $S$ may contain some information not present in format $T$,
the $S \rightarrow T$ transformation takes
\emph{both} the edited string $s \in S$ and a string $t \in T$ as arguments.
Inuitively, the lens extracts the data ``shared'' between the two
formats from $s$
and uses $t$ to supply any missing information.
% required by $T$ that is missing
%as it generates an output $t'$.  If $s$ is missing information, the transformation should
%extract the missing components from $t$.

Initially, we hypothesized that extending the synthesis algorithms
of Miltner \ETAL~\cite{optician} to operate over symmetric rather than
bijective lenses
would be relatively straightforward:  Simply replace the bijective
combinators with symmetric ones 
and search using similar heuristics.  However, 
this naive approach runs into two difficulties. \dpw{I got into the
weeds a little bit there. 
The goal was to set up the idea that there are ``two parts'' to a symmetric
lens transformation: the information from $s$ and the information from $t$.  In
a future paragraph in the intro we are going to explain that the central heuristic
involves minimizing the information from $t$.}
